---
title: "Airport Analysis"
author: "James Abrams & ChatGPT"
date: "2023-06-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

<section style="border-top: 1px solid black;"></section>

# Introduction

This analysis focuses on the {airport} departures dataset. The dataset contains information about the number of departures to different airports from {airport_name}. We will explore the top destinations and categorize them as domestic or international flights.  

#### **Questions to be answered**

1. Where are the most popular destinations for outbound flights?
2. What proportion of flights are international vs. domestic?
3. Which airlines are the most prominent?
4. What are the most popular times of day for flights to depart?  


<section style="border-top: 1px solid black;"></section>

#### **What are the most popular destinations for outgoing flights from {airport_name}?**

To determine the most popular destinations for outgoing flights from {airport_name}, we have analyzed the dataset and identified the top 10 destinations based on the number of departures.


<section style="border-top: 1px solid black;"></section>

```{r airports, echo=FALSE}

library(dplyr)
library(ggplot2)
library(knitr)

# Plotting function for bar charts.
plot_bar_chart <- function(data, x, y, lab_x, lab_y, title) {
  ggplot(data, aes(x = reorder({{ x }}, {{ y }}), y = {{ y }})) +
    geom_bar(stat = "identity", fill = "skyblue", width = 0.5) +
    labs(x = lab_x, y = lab_y) +
    ggtitle(title) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Load the {airport} departures data
departures_dd <- read.csv("{airport}_departures_dd.csv")

# Select the top 10 airports and reorder them by number of departures
top_airports <- departures_dd %>%
  select(-c(airline_name, top_airlines, dep_time)) %>%
  distinct() %>%
  select(airport_name, top_airports) %>%
  arrange(desc(top_airports)) %>%
  head(10) %>%
  as.data.frame()


# Create the bar chart with ordered bars

plot_bar_chart(top_airports, airport_name, top_airports,
               "Airport", "Number of Departures",
               "Top 10 Most Popular Airports Destinations")



```

<section style="border-top: 1px solid black;"></section>  
The bar chart above displays the top 10 most popular airport destinations based on the number of daily departures from {airport_name}. The airports are ranked by the number of departures, with the highest number at the top.

As we can see from the chart, the most popular destination for outgoing flights from {airport_name} is **`r top_airports$airport_name[1]`**, with a total of **`r top_airports$top_airports[1]`** departures.

<section style="border-top: 1px solid black;"></section>

#### **What is the proportion of Domestic flights compared to International flights leaving the airport?**

To understand the distribution of flights leaving the airport, we have analysed the proportion of domestic and international flights.


<section style="border-top: 1px solid black;"></section>

```{r pie chart, echo=FALSE}

library(kableExtra)

# Perform data manipulation
flight_proportions <- departures_dd %>%
  group_by(destination) %>%
  summarize(total = n()) %>%
  mutate(proportion = round(total / sum(total), 2))



# Render the table using kable
kable(flight_proportions, format = "html", caption = "Proportion of Domestic Vs International Flights") %>%
  kable_styling(bootstrap_options = "striped")


# Create the pie chart
ggplot(flight_proportions, aes(x = "", y = proportion, fill = destination)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = c("Domestic" = "turquoise", "International" = "pink")) +
  theme_void() +
  labs(fill = "Flight Type") +
  ggtitle("Proportion of Domestic vs International Flights") +
  geom_text(aes(label = paste0(round(proportion * 100), "%")), position = position_stack(vjust = 0.5))


```

<section style="border-top: 1px solid black;"></section>

The table above shows the proportion of domestic and international flights leaving the airport. It provides insights into the balance between domestic and international air traffic.

As we can see, **`r round(sum(flight_proportions[flight_proportions$destination == "Domestic", "proportion"]) * 100, 2)`%** of the flights leaving the airport are **domestic**, while **`r round(sum(flight_proportions[flight_proportions$destination == "International", "proportion"]) * 100, 2)`%** are **international**.



<section style="border-top: 1px solid black;"></section>


#### **What are the top 10 most popular airlines departing the airport?**

To analyze the popularity of airlines, we have selected the top 10 airlines based on the number of departures. The list is arranged in descending order, with the highest number of departures at the top.

<section style="border-top: 1px solid black;"></section>

```{r airlines, echo=FALSE}

# Select the top 10 airlines and reorder them by number of departures
top_airlines <- departures_dd %>%
  select(-c(airport_name, top_airports, destination, dep_time)) %>%
  filter(!is.na(airline_name)) %>%
  distinct() %>%
  select(airline_name, top_airlines) %>%
  arrange(desc(top_airlines)) %>%
  head(10) %>%
  as.data.frame()


# Create the bar chart with ordered bars
plot_bar_chart(top_airlines, airline_name, top_airlines,
               "Airline", "Number of Departures",
               "Top 10 Most Popular Airlines")



```

<section style="border-top: 1px solid black;"></section>

The bar chart above displays the top 10 most popular airlines based on the number of departures. It provides insights into the prominence of different airlines at {airport_name}.  

As we can see, the most popular airline based on the number of departures is **`r top_airlines$airline_name[1]`**, with **`r top_airlines$top_airlines[1]`** outgoing flights.

<section style="border-top: 1px solid black;"></section>

#### **What are the busiest times of day?**

To identify the most popular times of day for flights to depart, we have analyzed the dataset and categorized the departure times into hourly time slots.

<section style="border-top: 1px solid black;"></section>

```{r peak_hours, echo=FALSE}

library(lubridate)

# Use 24 hour clock format for this exercise
peak_hours <- departures_dd %>%
  select(dep_time) %>%
  mutate(dep_time = as.POSIXct(dep_time)) %>%
  mutate(timeslot = floor_date(dep_time, "hour")) %>%
  mutate(timeslot = format(timeslot, format = "%H")) %>%
  group_by(timeslot) %>%
  summarize(total = n()) %>%
  arrange(-desc(as.numeric(timeslot)))

# Convert timeslot to a factor with ordered levels
peak_hours$timeslot <- factor(peak_hours$timeslot, levels = unique(peak_hours$timeslot), ordered = TRUE)


# Generate a data frame with all hours
all_hours <- data.frame(timeslot = sprintf("%02d", 0:23))

# Merge all_hours with peak_hours
merged_data <- merge(all_hours, peak_hours, by = "timeslot", all.x = TRUE)

# Fill missing values with 0
merged_data$total[is.na(merged_data$total)] <- 0

# Convert timeslot to a factor with ordered levels
merged_data$timeslot <- factor(merged_data$timeslot, levels = unique(merged_data$timeslot), ordered = TRUE)

# Plot the time series bar chart with adjusted x-axis labels
ggplot(merged_data, aes(x = timeslot, y = total)) +
  geom_bar(stat = "identity", fill = "seagreen") +
  scale_x_discrete(breaks = merged_data$timeslot[seq(1, nrow(merged_data), by = 2)], 
                   labels = function(x) paste0(x, ":00")) +
  labs(x = "Timeslot", y = "Total") +
  ggtitle("Time Series Bar Chart Showing Airport Peak Hours by departures per hour") +
  theme_minimal()


```

<section style="border-top: 1px solid black;"></section>

The time series bar chart above illustrates the distribution of departures per hour, highlighting the busiest periods at the airport. Each bar represents a one-hour time slot, and the height of the bar represents the total number of departures during that time slot.

Based on the chart, the busiest time of day for departures from {airport_name} is **`r paste0(peak_hours$timeslot[which.max(peak_hours$total)], ":00")`**. The number of flights departing at this time is: **`r peak_hours$total[which.max(peak_hours$total)]`**. During this period, there is a significant surge in departures, indicating a high level of activity at the airport.

Conversely, the quietest time of day for departures is **`r paste0(peak_hours$timeslot[which.min(peak_hours$total)], ":00")`**. The number of flights departing at this time is: **`r peak_hours$total[which.min(peak_hours$total)]`**. During this period, the number of departures is relatively low, indicating a more peaceful and less crowded environment at the airport.

<section style="border-top: 1px solid black;"></section>

